---
title: "Balancing Utility and Privacy"
bibliography: ../references.bib
---

```{r}
#| include: false

library(sdcMicro)
library(tidyverse)

# Load data based on downloaded file
sdc_micro <- readRDS("../sdc_noise.rds") # Change based on the location of your Sdc-object file
data_withoutdirectidentifiers <- read.csv("../SimulatedData_noidentifiers.csv") # Change based on the location of your data file
```

As we have seen, anonymization techniques are all about hiding or obscuring the data or links within the data. This, of course, comes at a cost: The data may lose some of its **utility**.

We therefore need to find a well-calibrated balance between utility and anonymization, the famous sweet spot where we share data "as open as possible, as closed as necessary." To achieve this, we can utilize various measures - many of which we have already seen in [the chapter on assessing privacy risks](../2_ANONYMISATION_TECHNIQUES/2_1_Planning_Privacy) and in various `sdcMicro` outputs.

# Measuring Privacy

-   Calculate privacy levels (k-anonymity) and compare to the value measured before

*Include exercise on calculating the privacy level with k-anonymity like before in 2_1_Planning Privacy*

The most straightforward way to measure privacy after anonymization is to compare k-anonymity before and after. In [the chapter on assessing privacy risks](../2_ANONYMISATION_TECHNIQUES/2_1_Planning_Privacy), we found that nearly every participant in the original dataset had a unique combination of key variables. After applying non-perturbative techniques in [the chapter on non-perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_3_Non_Perturbative), we achieved 3-anonymity - every record shares its key variable combination with at least two others. Printing the sdcObject gives you the current k-anonymity level directly:

```{r}
print(sdc_micro, type = "risk")
```

The k-anonymity and disclosure risks are objective values, but always need to be considered in context: there are factors that are relevant here: what are key variables? what external data exist? what is the sampling strategy and what is known about that?

A less systematic approach to help here is red teaming data anonymization: Red teaming data anonymization [@Jansen2026_PuttingPrivacyTesta]

# Measuring Utility

How do we know whether our anonymized data is still useful? There are several approaches to measuring utility, ranging from simple comparisons to formal indices.

*Provide overview of utility measurements and maybe try out one*

-   utility indices [@Carvalho2023_SurveyPrivacyPreservingTechniques]
    -   predictive performance measures for machine learning *(potentially as call-out box)*
    -   information loss measures:
        -   distance/distribution comparisons
        -   a penalty of transformations through generalisation and suppression
        -   statistical differences
-   assessing utility for the current use case
    -   perform statistical analysis before and after anonymization
    -   see if results are comparable

*Include exercise for calculating information loss measure*

`sdcMicro` provides two built-in information loss measures for numeric variables. **IL1** is the average absolute difference between original and anonymized values, relative to the original. A value of 0 means no change; higher values mean more distortion. The **eigenvalue ratio** compares the covariance structure of the original and anonymized data - values close to 1 indicate that the relationships between variables have been preserved. Both were already shown in the sdcMicro output after applying perturbative techniques in [the chapter on perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_4_Perturbative):

```{r}
print(sdc_micro, type = "risk")
```

For non-numeric variables that were recoded or suppressed, there is no single formal measure - but you can compare frequency tables before and after. For example, comparing the distribution of age bands or religion categories tells you whether the recoding preserved the overall composition of the sample.

Beyond these built-in measures, the most practically relevant utility check is simply to re-run the analysis that motivated data collection on both the original and anonymized datasets and compare the results. For our dataset, that means checking whether the association between religion and political opinion is still visible after anonymization. If the key result holds, the anonymized data is fit for purpose.

# Striking the Right Balance

-   iterative process: rework anonymization after measuring both

Balancing privacy and utility is not a one-shot decision - it is an iterative process. After measuring both, you may find that your current anonymization is too aggressive (data is well-protected but the key analysis no longer works) or too lenient (utility is preserved but too many records are still unique). In either case, you go back to the anonymization steps and adjust: changing recoding thresholds, relaxing or tightening the k-anonymity target, or choosing a different perturbative technique.

For our dataset, a reasonable workflow looks like this: start by checking the k-anonymity level and disclosure risk on `sdc_nonpert` (from [the chapter on non-perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_3_Non_Perturbative)), then check the information loss after applying perturbative techniques on `sdc_micro` or `sdc_noise` (from [the chapter on perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_4_Perturbative)), and finally compare the religion-political opinion analysis on original and anonymized data. If the analysis still holds and k-anonymity is at least 3, the anonymization is likely sufficient for sharing.

There are no universal thresholds for "good enough" privacy or utility. A commonly used rule of thumb is that k-anonymity should be at least 3-5 for most research datasets, meaning every combination of quasi-identifiers appears at least 3-5 times. For utility, IL1 values close to 0 and eigenvalue ratios close to 1 indicate low distortion. But ultimately, the right balance depends on your specific context: how sensitive is the data, who will have access, and what analyses need to be supported. When in doubt, consult your institution's data protection officer or data steward.

*Link to people to contact at LMU (data stewards? open science team?)*

*Include finalization of data (export)*

# Exercise

Evaluate the utility of the dataset using `sdcMicro` functions.

After anonymization, you need to check that the data is still **useful** - that the patterns you care about (e.g., the relationship between religion and political opinion) have not been destroyed. `sdcMicro` provides utility measures, but it is also worth comparing distributions and running the actual analysis on both the original and anonymised data.

Key utility concepts:

-   **IL1** — average absolute difference between original and anonymised numeric values (lower = better)
-   **Eigen** — compares covariance structure; values close to 1 mean the relationships between variables are preserved
-   **`utility()` / `print(..., type = "utility")`** — overall summary
-   **Visual comparison** — plotting distributions side-by-side is often the most intuitive check

## Exercise: Evaluating Data Utility

Use the `sdc_micro` results from the previous exercises to compare it to the original data.

1.  Print the utility summary of `sdc_micro`. What do IL1 and the eigen measure tell you?

2.  Compare the income distribution before and after anonymisation. Plot histograms of the original and anonymised income side-by-side.

3.  Re-run the analysis that motivated data collection**:** compute the mean political opinion scores (`pol_immigration`, `pol_redistribution`) by religion group on both the original and anonymised data. Are the group differences preserved?

4.  Reflect**:** based on your findings, is the anonymised dataset fit for purpose? Would you adjust any of the anonymisation steps?

::: {.callout-note collapse="true"}
## Solution

### Step 1 – Utility summary

```{r}
sdc_micro
```

IL1 close to 0 indicates little numeric distortion. Eigen values close to 1 indicate that the covariance structure between numeric variables is preserved. If you applied strong microaggregation (small `k` → large groups), you will see higher IL1 and eigen values further from 1.

### Step 2 – Compare income distributions

```{r}
library(ggplot2)

# Extract anonymised income from the sdc object
income_original   <- data_withoutdirectidentifiers$income
income_anonymised <- get.sdcMicroObj(sdc_micro, "manipNumVars")$income

income_compare <- tibble(
  income = c(income_original, income_anonymised),
  version = rep(c("Original", "Anonymised"), each = length(income_original))
)

ggplot(income_compare, aes(x = income, fill = version)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  scale_fill_manual(values = c("Original" = "#2E4057", "Anonymised" = "#E07A5F")) +
  labs(
    title = "Income distribution: original vs. anonymised",
    x = "Annual income (€)", y = "Count", fill = NULL
  ) +
  theme_minimal()
```

The shapes should look similar. A heavily chunked distribution (tall bars at regular intervals) would indicate that microaggregation groups were too large.

*Check data: anonymized data virtually identical with non-anonymized*

### Step 3 – Check whether the key analysis still works

```{r}
# Analysis on original data
original_summary <- data_withoutdirectidentifiers %>%
  group_by(religion) %>%
  summarise(
    mean_immigration    = mean(pol_immigration,    na.rm = TRUE),
    mean_redistribution = mean(pol_redistribution, na.rm = TRUE),
    n = n()
  )

# Rebuild anonymised dataset with the manipulated variables
anon_data <- get.sdcMicroObj(sdc_micro, "manipKeyVars") %>%
  bind_cols(get.sdcMicroObj(sdc_micro, "manipNumVars")) %>%
  bind_cols(
    data_withoutdirectidentifiers %>% select(religion, starts_with("pol_"))
  )

anonymised_summary <- anon_data %>%
  group_by(religion) %>%
  summarise(
    mean_immigration    = mean(pol_immigration,    na.rm = TRUE),
    mean_redistribution = mean(pol_redistribution, na.rm = TRUE),
    n = n()
  )

original_summary
anonymised_summary
```

If the group means are close between both summaries, the anonymisation has preserved the analytical signal. Political opinion variables were not altered (they are neither key variables nor numeric variables in the sdcMicro object), so differences here would only arise if the grouping structure changed dramatically.

### Step 4 – Reflection

There is no single correct answer. Consider:

-   If income distributions look very similar and group means are preserved → the anonymisation is fit for purpose for this analysis
-   If you used `k = 5` and results still look good, increasing to `k = 3` would reduce distortion while still providing some protection, but weigh this against the risk reduction you found in the earlier exercises
-   Variables like `plz_region` and `age_group` were generalized; if a future analysis requires finer geographic or age breakdowns, you would need to rework the anonymization strategy
:::

Let's keep in mind, that while there is a tension between data protection and openness, in practice, these principles enable each other: Sharing data builds trust in research findings, while protecting participants builds trust in researchers. When done well, anonymization is the bridge between these goals - it lets you share data openly while honoring the privacy expectations of the people who provided it [@Jansen2025_TensionOpenData].

#### Learning Objective

-   After completing this part of the tutorial, you will be able to make informed decisions when balancing the risks and utility of the anonymized data.

#### Exercises

-   apply measurement of k-anonymity

-   apply measurement of utility

#### Resources, Links, Examples

-   examples in UK: <https://ukdataservice.ac.uk/deposit-data/sharing-experiences/>

# References
