---
title: "Balancing Utility and Privacy"
bibliography: ../references.bib
---

::: {.callout-tip collapse="true"}
## Learning Objective

-   After completing this part of the tutorial, you will be able to make informed decisions when balancing the risks and utility of the anonymized data.
:::

```{r}
#| include: false

# Setup: load packages and the saved sdcObjects
library(sdcMicro)
library(tidyverse)

sdc_nonpert <- readRDS(here::here("sdc_nonpert.rds"))
sdc_noise <- readRDS(here::here("sdc_noise.rds")) # Change based on the location of your Sdc-object file
sdc_micro <- readRDS(here::here("sdc_micro.rds"))
data_withoutdirectidentifiers <- read.csv(here::here("SimulatedData_noidentifiers.csv")) # Change based on the location of your data file
```

As we have seen, anonymization techniques are all about hiding or obscuring the data or links within the data. This, of course, comes at a cost: The data may lose some of its **utility**.

We therefore need to find a well-calibrated balance between utility and anonymization, the famous sweet spot where we share data "as open as possible, as closed as necessary." To achieve this, we can utilize various measures—many of which we have already seen in [the chapter on assessing privacy risks](../2_ANONYMISATION_TECHNIQUES/2_1_Planning_Privacy) and in various `sdcMicro` outputs.

# Measuring Privacy

The most straightforward way to measure privacy after anonymization is to compare k-anonymity before and after applying an anonimization technique. In [the chapter on assessing privacy risks](../2_ANONYMISATION_TECHNIQUES/2_1_Planning_Privacy), we found that nearly every participant in the original dataset had a unique combination of key variables. After applying non-perturbative techniques in [the chapter on non-perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_3_Non_Perturbative), we achieved 3-anonymity—every record shares its key variable combination with at least two others. Printing the `sdcObject` gives you the current k-anonymity level directly:

```{r}
# Check the current k-anonymity level
sdc_nonpert
```

The k-anonymity and disclosure risks are objective values, but always need to be considered in context: What variables did we define within the `sdcObject`? What external data exist? What is the sampling strategy and what is known about that?

In practice, it can be challenging to correctly assess these risks given the complexity of research projects and existing information. In such cases, red teaming the data anonymization can help [@Jansen2026_PuttingPrivacyTesta]: Another person (e.g., another member of your lab) is instructed to play the role of an attacker, by, e.g., searching for external information or singling out the least protected individuals.

In general, it is worth it to check the assumptions of the anonymization process every few years after publication of the data, especially if new techniques have emerged (see callout box below).

::: {.callout-note collapse="false"}
## Deanonymization Using AI

AI tools are increasingly being used to **re-identify individuals from datasets that were considered anonymous**. Recent research shows that large language models (LLMs) can match anonymous text profiles to named individuals across platforms—for example, linking an anonymous forum post to a person's identified social media account [@lermen]. Separately, research has shown that individuals can be re-identified from behavioral traces alone, including web browsing history, geolocation data, and even face scans, by combining datasets that each seem innocuous on their own [@rocher2025].

For **quantitative survey** or experimental data of the kind covered in this tutorial, I do not expect AI tools to fundamentally change the re-identification risk—the attack surface is just not the same as for text or behavioral data. That said, AI can make an **attacker more efficient**: tasks like searching for external information, generating plausible attack combinations, or automating record linkage are all easier with modern AI tools.
:::

# Measuring Utility

How do we know whether our anonymized data is still useful? There are several approaches to **measuring utility**, ranging from simple comparisons to formal indices.

`sdcMicro` provides two built-in information loss measures for numeric variables. **IL1** is the sum of absolute differences between original and anonymized values, relative to the original. A value of 0 means no change; higher values mean more distortion. The **eigenvalue ratio** compares the eigenvalues of the original covariance matrix to the eigenvalues of the anonymized data's covariance matrix. Values close to 0 indicate the least changes. These indicators are primarily useful to compare different anonymization methods with each other.

Both values are part of the standard `sdcMicro` output after applying any anonymization technique.

# Exercise: Evaluating Data Utility

## Using `sdcMicro`'s Parameters

Let's go back to the two `sdcObjects` we created in [the chapter on perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_4_Perturbative): `sdc_micro` (where we applied microaggregation) and `sdc_noise` (where we applied noise).

1.  Compare the utility with IL1 and the difference of eigenvalues. Which method do you prefer when looking at these utility indices?
2.  Make a decision for one of the methods and export the data for further use.

::: {.callout-important collapse="true"}
## Solution

Inspect the `sdcObjects` to see the values of IL1 and the eigenvalue difference:

```{r}
# Compare utility: IL1 and eigenvalue difference
sdc_micro
sdc_noise
```

Both indices prefer the microaggregation approach: The IL1 is much smaller, indicating that values are closer to the original data, and the difference of eigenvalues is slightly smaller, also indicating that microaggregation preserves the most utility.

Now I can extract the dataset from the `sdc_micro` object. Because `sdcMicro` stores recoded key variables as integer codes internally, I need to restore the labels manually using `mutate`:

```{r}
# Extract the anonymized data and restore category labels
data_anonymized <- extractManipData(sdc_micro)

# Restore labels for variables recoded inside sdcMicro
age_labels    <- c("18-29", "30-39", "40-49", "50-59", "60+")
plz_labels    <- c("8xxxx", "other")

data_anonymized <- data_anonymized %>%
  mutate(
    age = factor(age, levels = seq_along(age_labels), labels = age_labels),
    plz = factor(plz, levels = seq_along(plz_labels), labels = plz_labels)
  )

```
:::

## Using Statistics

For non-numeric variables that were recoded or suppressed, there is **no single formal measure**—but you can compare frequency tables before and after. For example, comparing the distribution of age bands or religion categories tells you whether the recoding preserved the overall composition of the sample.

Beyond these built-in measures, the most **practically relevant utility check is simply to re-run the analysis** that motivated data collection on both the original and anonymized datasets and compare the results. For our dataset, that means checking whether the association between religion and political opinion is similar to before anonymization. If the key result holds, the anonymized data is fit for purpose.

1.  Compare descriptive statistics on the indirect identifiers age, gender, education, income, plz, and years in job before and after anonymization.
2.  Compare the relation between religion and the political opinion variables `pol_immigration`, `pol_environment`, `pol_redistribution`, and `pol_eu_integration` before and after anonymization.
3.  Reflect on these comparisons: Is utility well-enough preserved?

::: {.callout-important collapse="true"}
## Solution

**Descriptive statistics on indirect identifiers**

Compare distributions before and after using `summary()` and frequency tables:

```{r}
# Compare descriptive statistics before and after anonymization
# Before anonymization
summary(data_withoutdirectidentifiers[, c("age", "income", "years_in_job")])
table(data_withoutdirectidentifiers$gender, useNA = "always") # Make sure to show NAs as well
table(data_withoutdirectidentifiers$education, useNA = "always")
table(data_withoutdirectidentifiers$plz, useNA = "always")

# After anonymization
summary(data_anonymized[, c("income", "years_in_job")])
table(data_anonymized$age, useNA = "always")
table(data_anonymized$gender, useNA = "always")
table(data_anonymized$education, useNA = "always")
table(data_anonymized$plz, useNA = "always")
```

Age and postal code are now categorical bands, so the numeric summary is no longer meaningful—use frequency tables instead. Income and years in job are still numeric after microaggregation; their medians are very close to the original, but especially income's mean is lower now—this is due to the extreme outliers that were top-coded.

For gender, 6 non-binary individuals' data was suppressed to achieve necessary k-anonymity.

For education, all individuals without a degree were recoded to NA; in each other category, few individuals's data was suppressed. This might be a problem for data utility, especially if our goal was to show that we reached individuals from all educational levels. If this was the case, I'd recommend to go back to applying local suppression and to prioritize that educational level stays non-suppressed.

We lost, as intended, a lot of information on plz and age. For our use case, this is okay.

**Religion and political opinions**

```{r}
# Compare religion vs. political opinion before and after anonymization
# Before
data_withoutdirectidentifiers %>%
  group_by(religion) %>%
  summarise(across(c(pol_immigration, pol_environment, pol_redistribution, pol_eu_integration), mean, .names = "mean_{.col}"))

# After
data_anonymized %>%
  group_by(religion) %>%
  summarise(across(c(pol_immigration, pol_environment, pol_redistribution, pol_eu_integration), mean, .names = "mean_{.col}"))
```

The group-level means should be very the same between the original and anonymized datasets. The key research question—whether religion predicts political opinion—is still answerable. We did not change anything regarding these variables.

**Reflection**

For the purpose of this dataset—studying the relationship between religion and political opinion—utility is well-preserved. The main variable of interest (religion) is still present, just with fewer fine-grained categories. The political opinion variables are unchanged. Income and years in job show very similar distributions after microaggregation.
:::

# Striking the Right Balance

Balancing privacy and utility is not a one-shot decision—it is an **iterative process**. After measuring both, you may find that your current anonymization is too aggressive (data is well-protected but the key analysis no longer works) or too lenient (utility is preserved but too many records are still unique). In either case, you go back to the anonymization steps and adjust: changing recoding thresholds, relaxing or tightening the k-anonymity target, or choosing a different perturbative technique.

For our dataset, a reasonable workflow looks like this: start by checking the k-anonymity level and disclosure risk on `sdc_nonpert` (from [the chapter on non-perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_3_Non_Perturbative)), then check the information loss after applying perturbative techniques on `sdc_micro` or `sdc_noise` (from [the chapter on perturbative techniques](../2_ANONYMISATION_TECHNIQUES/2_4_Perturbative)), and finally compare the religion-political opinion analysis on original and anonymized data. If the analysis still holds and k-anonymity is at least 3, the anonymization is likely sufficient for sharing.

There are **no universal thresholds** for "good enough" privacy or utility. A commonly used rule of thumb is that k-anonymity should be at least 3-5 for most research datasets, meaning every combination of indirect identifiers appears at least 3-5 times (see discussion above). For utility, IL1 values and eigenvalue differences close to 0 indicate low distortion. But ultimately, the right balance depends on your specific context: how sensitive is the data, who will have access, and what analyses need to be supported. When in doubt, consult your institution's data protection officer ([LMU Munich data protection office](https://www.lmu.de/en/about-lmu/structure/organizational-structure/officers-representatives-and-contact-persons/data-protection-officer.html)), data steward (or a local research data management team, [LMU Munich research data management team](https://www.ub.lmu.de/en/open-access-publishing/research-data/)), or open science team ([LMU Open Science Center](https://www.osc.lmu.de/)).

::: {.callout-note collapse="true"}
## Who Is Responsible If Someone Is Re-Identified Anyway?

*Disclaimer: We do not represent LMU Munich, and we do not have the authority to give legal advice or to approve your anonymization. What follows is our personal reading of the legal situation, intended as orientation—for binding answers, contact your institution's data protection officer.*

You might ask yourself: "What if I do everything right, and someone still gets re-identified? Am I liable?" We don't think you are responsible for two reasons:

First, the **legal standard is not zero risk**. Under the GDPR ([Recital 26](https://gdpr-info.eu/recitals/no-26/)), data counts as anonymous if re-identification is not *reasonably likely*, considering cost, time, and available technology. Absolute anonymization is usually neither possible nor required—it is sufficient that re-identification is practically infeasible or requires disproportionate effort. The relevant question is therefore not "Could someone ever be re-identified?" but "**Was the anonymization decision reasonable at the time?**". A later re-identification does not retroactively prove your procedure was negligent.

Second, **legally, who is responsible?** Not you personally, in most cases: the *controller* under the GDPR is typically your institution (e.g., the university), not the individual researcher. The data protection officer advises and monitors compliance—they do not carry the responsibility alone either, and their approval is not a magic shield. What protects you is a documented, institutionally backed process. **You are safe if you followed:**

1.  a written anonymization plan, agreed on before release;
2.  the advice of your data protection officer (and, where required, a data protection impact assessment);
3.  a documented risk assessment (attack scenarios, indirect identifiers, external linkage, a k-anonymity target or comparable metric—keeping in mind that no k-value is a magic legal threshold);
4.  a deliberate release decision (fully open, restricted access, or another access model);
5.  and did not deviate from the approved plan.

If someone is re-identified anyway, that is an institutional matter—reassessing the dataset, restricting access, notifying where required—not evidence of your personal fault. In our **personal opinion**, researchers who follow an approved, documented procedure have fulfilled their obligations and are not at fault. The circular logic of "someone was re-identified, therefore your anonymization was too lenient" gets it backward.
:::

Let's keep in mind that, while there is a **tension between data protection and openness**, in practice, these principles **enable** each other: Sharing data builds trust in research findings, while protecting participants builds trust in researchers. When done well, **anonymization is the bridge** between these goals—it lets you share data openly while honoring the privacy expectations of the people who provided it [@Jansen2025_TensionOpenData].

#### Resources, Links, Examples

-   [Examples of decisions for balancing utility and privacy from the UK](https://ukdataservice.ac.uk/deposit-data/sharing-experiences/).

# References
