---
title: "Welcome to the Data Anonymization Course!"
---

## Tutorial Overview

This **self-paced tutorial** on anonymization of quantitative research data is intended to take about three hours to complete.

The tutorial is split into the following sections:

1.  [**FOUNDATIONS OF DATA PROTECTION**](foundations/introduction-data-privacy) talks about data protection basics in ethics and law, mechanisms of data protection in research, and basic terms.
2.  [**DATA ANONYMIZATION PROCESS**](anonymisation-techniques/planning-privacy) walks you through the process of anonymizing your research data based on example data.
3.  [**BALANCING DATA PROTECTION AND OPENNESS**](privacy-openness/balancing) presents methods for aligning your data protection and open science interests.
4.  [**ANONYMIZATION WORKFLOW**](workflow/summary) closes this tutorial by summarizing the learned workflow.

## What You'll Learn

By the end of this tutorial, you will be able to:

-   **Understand key concepts** in the world of privacy (e.g., anonymization, k-anonymity)
-   **Classify data** in relevant categories for data protection (e.g., personal data, sensitive data)
-   **Apply anonymization techniques** using R in a coherent workflow
-   **Make informed decisions when balancing the risks and utility** of the anonymized data

## What You Will **NOT** Learn

You will not learn anything other than anonymization of quantitative tabular data.

Here are a few helpful links for other data types:

-   Anonymizing neuroimaging metadata ([article](https://doi.org/10.1162/IMAG.a.1144))
-   Anonymizing qualitative textual data with the tool QualiAnon ([video tutorial](https://www.youtube.com/watch?v=RYQn4DjdKmo))
-   Anonymizing sensitive qualitative data ([lecture recording](https://www.youtube.com/watch?v=-Pha3EINF3s&ab_channel=CenterforOpenScience) and [article](https:/doi.org/10.1037/tra0001358))

## Prerequisites

You need basic R skills (e.g., loading data and packages). Experience with data wrangling with `tidyverse` is beneficial.

## Required Software

To follow along with the hands-on exercises, you will need a recent version of [R](https://cran.r-project.org/) (version 4.1.0 or newer) and [RStudio](https://posit.co/download/rstudio-desktop/).

You will also need a handful of R packages. Install them with:

```{r}
#| eval: false
install.packages("sdcMicro")  # measure re-identification risk and apply anonymization techniques
install.packages("tidyverse") # data wrangling and plotting, used throughout the exercises
install.packages("here")      # locate your data file regardless of the working directory
install.packages("knitr")     # render tables in the tutorial's examples
install.packages("xfun")      # small helpers used in the tutorial's examples
```

## The Example Dataset

Throughout the tutorial, we work with the same **simulated dataset** of 200 Germans. The data's purpose is to answer whether **certain political opinions are linked to religion**. We will assess its privacy risks and then anonymize it step by step across the exercises. Download it here:

```{r}
#| label: download-link
#| echo: false
# Generates a clickable download link for the GitHub Pages website
xfun::embed_file(
  here::here("simulated-data.csv"),
  text = "⬇️ Download simulated-data.csv"
)
```

Once downloaded, copy this into a new Markdown file in RStudio to import the data for further analysis:

```{r}
#| label: data-import
#| message: false

# Load data based on downloaded file
data <- read.csv(here::here("simulated-data.csv")) # Change based on the location of your data file

```

The dataset contains the following variables:

::: {.callout-note collapse="true"}
### Data Dictionary

| Variable Name | Description | Item | Values |
|----|----|----|----|
| id | Number assigned to each participant in order of participation | (assigned in background) | Integer; 1-200 |
| name | First and last name of participant | Please indicate your full name (first and last name) | String of characters |
| email | Email address of participant | What is your e-mail address? | String of characters |
| plz | German postal code | What is your postal code? | String of characters |
| gender | Gender of participant | What is your gender? | Factor; "male"/"female"/"non-binary" |
| age | Age of participant in years | What is your age in years? | Integer; 18-100 |
| income | Personal annual income in Euros | What was your income over the last twelve months | Integer |
| religion | Religion of participant | What is your religion? | Factor; "Catholicism","Protestantism","Islam","Eastern Orthodoxy","Judaism", "Buddhism", "Hinduism", "Other", "None" |
| education | Highest degree of education | What is your highest degree of education? | Factor; "no degree","trade school","high school", "university","doctoral title" |
| pol_immigration | Likert item measuring opinion on immigration | The government should limit immigration more strictly than it currently does. | Integer; 1-5 |
| pol_environment | Likert item measuring opinion on environment | Protecting the environment should be a top priority, even if it slows economic growth. | Integer; 1-5 |
| pol_redistribution | Likert item measuring opinion on redistribution of wealth | The government should reduce income differences between rich and poor. | Integer; 1-5 |
| pol_eu_integration | Likert item measuring opinion on membership in EU | Our country benefits from being a member of the European Union. | Integer; 1-5 |
| ip_address | IP address (version 4) of participant's device when answering survey | (collected in background) | String of characters |
| years_in_job | Number of years the participant has been in their current job | How many years have you been in your current job? | Integer; 0-n |
:::

:::: {.columns}

::: {.column width="47.5%"}

::: {.callout-tip title="Support the LMU Open Science Center!"}

**Like this tutorial? Be sure to give [the repository]({{< meta repo-url >}}) a star and [follow us on GitHub](https://github.com/lmu-osc)!**
:::

:::

::: {.column width="5%"}
:::

::: {.column width="47.5%"}


::: {.callout-tip title="Questions about this tutorial?"}
**If you have questions for the author of this tutorial, please take a look at the [About](about.qmd) page.**
:::

:::

::::
