---
title: "Planning Data Protection"
bibliography: ../references.bib
---

Assess the privacy risk by applying k-anonymity with sdcMicro to example data

The sdcMicro R package

sdcMicro is an R package specialized in the anonymisation of micro (small) data sets. sdcMicro offers a command line and a Graphical User interface (GUI).

Let's generate some fake data: (*change when exercises are finalized, I'll probably use test data provided by the package)*

```{r}
data <- data.frame(
  age = c(34, 35, 36, 36, 36, 40, 40, 41, 42, 42),
  gender = c("M", "M", "M", "F", "F", "M", "F", "F", "M", "M"),
  zipcode = c(
    "12345",
    "12345",
    "12345",
    "12345",
    "12345",
    "67890",
    "67890",
    "67890",
    "67890",
    "67890"
  ),
  diagnosis = c(
    "Flu",
    "Cold",
    "Flu",
    "Cold",
    "Allergy",
    "HIV",
    "HIV",
    "Flu",
    "Cold",
    "Allergy"
  )
)
```

```{r}
 #| echo: false   
 knitr::kable(head(data), caption = "Raw data example")
```


#### 

#### Learning Objective

-   After completing this part of the tutorial, you will be able to calculate individual privacy risks for continuous variables by utilizing k-anonymity.

#### Exercises

-   calculate the privacy level for certain variables

#### Resources, Links, Examples

The sdcMicro R package

sdcMicro is an R package specialized in the anonymisation of micro (small) data sets. sdcMicro offers a command line and a Graphical User interface (GUI).

Let's generate some fake data: (*change when exercises are finalized, I'll probably use test data provided by the package)*

```{r}
data <- data.frame(
  age = c(34, 35, 36, 36, 36, 40, 40, 41, 42, 42),
  gender = c("M", "M", "M", "F", "F", "M", "F", "F", "M", "M"),
  zipcode = c(
    "12345",
    "12345",
    "12345",
    "12345",
    "12345",
    "67890",
    "67890",
    "67890",
    "67890",
    "67890"
  ),
  diagnosis = c(
    "Flu",
    "Cold",
    "Flu",
    "Cold",
    "Allergy",
    "HIV",
    "HIV",
    "Flu",
    "Cold",
    "Allergy"
  )
)
```

```{r}
#| echo: false  
knitr::kable(head(data), caption = "Raw data example")
```

#### To Do List

-   Decide on federated learning
