---
title: "Planning Anonymization"
bibliography: ../references.bib
---

*Explain a bit about where we are in the workflow of research now*

Anonymization of data, in most cases, means covering or obscuring the data, meaning there is always a loss of information. However, when sharing data, we need to keep their utility; otherwise, they are rendered useless.

The goal is to use the appropriate anonymization techniques to achieve an acceptable level of risk for each individual, while at the same time, keeping utility as high as possible (see also [BALANCING DATA PROTECTION AND OPENNESS](../3_PRIVACY_OPENNESS/3_1_Balancing)).

To achieve that, before starting the anonymization process, we will assess the level of privacy risk in our data and investigate potential risks, taking what we learned in the [last chapter](../1_FOUNDATIONS/1_5_Privacy_Concepts) and putting it into practice.

# The `sdcMicro` Package

To that end (and for most of the following exercises), we will use the package `sdcMicro` in R and the test dataset it contains. You already learned about the data it contained in the last exercise (*add link to exercise*)

sdcMicro is an R package specialized in the anonymisation of micro (small) data sets. sdcMicro offers a command line and a Graphical User interface (GUI).

Let's generate some fake data: (*change when exercises are finalized, I'll probably use test data provided by the package)*

```{r}
data <- data.frame(
  age = c(34, 35, 36, 36, 36, 40, 40, 41, 42, 42),
  gender = c("M", "M", "M", "F", "F", "M", "F", "F", "M", "M"),
  zipcode = c(
    "12345",
    "12345",
    "12345",
    "12345",
    "12345",
    "67890",
    "67890",
    "67890",
    "67890",
    "67890"
  ),
  diagnosis = c(
    "Flu",
    "Cold",
    "Flu",
    "Cold",
    "Allergy",
    "HIV",
    "HIV",
    "Flu",
    "Cold",
    "Allergy"
  )
)
```

```{r}
 #| echo: false   
 knitr::kable(head(data), caption = "Raw data example")
```

# Identifying Privacy Risks

*Include exercising on identifying possible risks within the data*

*Include examples for attack scenarios*

# Measuring Privacy Risks

*Explain process of assessing the risk with sdcMicro*

*Include exercise on applying this to the example data*

At this point, you decide whether the level of privacy risks is acceptable. If so, you do not need anonymization but can skip to preparing your data for publication.

#### Learning Objective

-   After completing this part of the tutorial, you will be able to calculate individual privacy risks for continuous variables by utilizing k-anonymity.

#### Exercises

-   calculate the privacy level for certain variables

#### Resources, Links, Examples

#### To Do List

-   Decide on federated learning
