---
title: "Assessing Privacy Risks"
bibliography: ../references.bib
---

```{r Setup}
#| include: false

data_withoutdirectidentifiers <- read.csv("../SimulatedData_noidentifiers.csv") # Change based on the location of your data file
```

In the previous chapters, we covered the foundations: what counts as personal data, what the GDPR requires, what anonymization mechanisms exist, and what privacy risks look like. Now we move to the practical part - actually anonymizing data. But before we jump into specific techniques, we first need to assess how much risk our data carries. That is what this chapter is about.

Anonymization of data, in most cases, means covering or obscuring the data, meaning there is always a loss of information. However, when sharing data, we need to keep their utility; otherwise, they are rendered useless.

The goal is to use the appropriate anonymization techniques to achieve an acceptable level of risk for each individual, while at the same time, keeping utility as high as possible (see also [BALANCING DATA PROTECTION AND OPENNESS](../3_PRIVACY_OPENNESS/3_1_Balancing)).

To achieve that, before starting the anonymization process, we will assess the level of privacy risk in our data and investigate potential risks, taking what we learned in the [last chapter](../1_FOUNDATIONS/1_5_Privacy_Concepts) and putting it into practice.

# Identifying Privacy Risks

Take another look at the dataset from [Chapter 1.3](../1_FOUNDATION/1_3_Personal_Data), the version without direct identifiers. Try to answer the following questions:

-   Which combinations of indirect identifiers could potentially single out an individual?
-   What attack scenarios are plausible? Think about who might try to re-identify someone in this dataset and what external information they might have access to.
-   What type of disclosure is most likely for this dataset - identity disclosure, attribute disclosure, or membership disclosure?

::: {.callout-note collapse="true"}
## Solution

*Add concrete combinations*

Here are a few concrete attack scenarios to consider, based on our dataset of 200 individuals in Germany (with variables: gender, age, education, plz, income, religion, job_title, years_in_job, and political opinion variables):

-   A colleague knows that a specific person lives in a particular postal code area, works as a "Data Scientist," and is 34 years old. If the combination of plz, job_title, and age is unique in the dataset, that colleague can identify the person and look up their income and political opinions. This is identity disclosure. In this scenario, the attacker has the knowledge that his colleague has participated in the study. In practice, when calculating risks, we usually assume no such knowledge (see callout box below).
-   A journalist cross-references the dataset with publicly available professional directories or LinkedIn profiles. By matching job_title, age, and education level, they could narrow down candidates and potentially identify individuals - especially in smaller postal code areas where few people share the same profession.
-   A neighbor knows that someone in their building participated in the study. They know the person's approximate age, gender, and religion. Even if they cannot pinpoint the exact row, they might learn something new about that person's income or political views. This is attribute disclosure - the attacker does not need to find the exact record, just a group of records that all share the same sensitive value.

These are just a few examples; in general, the existence and likelihood of risks is very much dependent on the variables in the dataset, the sensibility of data, the recruitment strategy, and many other factors. *Elaborate more/link to other relevant parts*
:::

*Add callout box whether participation knowledge needs to be assumed or not (and its implications)*

::: callout-note
Felix's notes:

-   "reasonably likely”: We should discuss here or somewhere else one of the main factors: The knowledge of an attacker whether a person is in a specific data set ("Teilnahmekenntnis"). This changes a lot for the k-anonymity. Here a my notes regarding this:

    -   The relevant keyword is "Teilnahmekenntnis" / "participation knowledge": The re-identification risk rises very much when attacks know who participated in a survey. Von Klaus Pforr ([https://sciences.social/\@klauspforr/110775060196195399](https://sciences.social/@klauspforr/110775060196195399)): "Never post your survey participation in a run-of-the-mill survey on social media because the whole anonymisation concept most of time assumes negligble prior probs for these. Whenever you see this in the wild, send a direct message that people should refrain from this behavior in the future"

    -   I talked to Johannes Breuer (who was at GESIS back then), and he said that the big surveys always assume NO Teilnahmekenntnis when evaluating their level of anonymity.
:::

# Measuring Privacy Risks

## The `sdcMicro` Package

To formally measure privacy risks (and for most of the following exercises), we will use the package `sdcMicro` in R.

sdcMicro is an R package specialized in the anonymization of micro (small) data sets. sdcMicro offers a command line and a Graphical User interface (GUI).

*Include more info on functions in sdcMicro?*

## Using `sdcMicro` To Assess Privacy Risks

Let's start by creating a sdcMicro object using the class `sdcMicroObj`. The object becomes the working environment for all subsequent anonymization and risk assessment steps. It stores

-   the original dataset,
-   the roles of variables within the dataset,
-   risk measures,
-   anonymized versions of the variables,
-   utility measures, and
-   the history of modifications.

`sdcMicro` updates risk and utilty estimates after each anonymization step.

For this object, we start with defining the following:

-   `dat`: the dataset we are working with

-   `keyVars`: indirect identifiers that are categorial (i.e., have categories) we want to assess the risk of (e.g., age, gender)

-   `numVars`: continuous (i.e., numeric) indirect identifiers (e.g., income)

`sdcMicro` needs, at the least, the data and categorial indirect identifiers to assess risks. We can later add more information to the object.

::: callout-note
## More Functions of sdcMicro

-   `extractManipData(sdc)`\
    Extracts the anonymized dataset.

-   `undolast(sdc)`\
    Reverts the last anonymization step.

-   `get.sdcMicroObj()`\
    Accesses internal components of the object.

*Add info on GUI usage*
:::

## Exercise: Assessing Privacy Risks With `sdcMicro`

1.  Create an object within sdcMicro. Inspect the object.
2.  What k-anonymity level is currently reached?

::: {.callout-note collapse="true"}
## Solution

Create the object. Provide gender, age, education, postal code, and years in job as categorial variables of interest (`keyVars`). The only numerical variable we are interested in is income (`numVars`).

```{r}
library(sdcMicro)

sdc_data <- createSdcObj(
  dat = data_withoutdirectidentifiers, 
  keyVars = c("gender", "age", "education", "plz", "years_in_job"),
  numVars = "income"
  )  

sdc_data # Summarizes the risks overall
```

`sdcMicro` shows us the unmodified, original parameters in parentheses. Right now, they are, of course, identical to the current parameters as we have not made any changes yet.

For the **categorial key variables**, we get information on the numbers and sizes of categories. The more categories and the smaller the number of entries in each category, the easier the de-identification as we have many unique individual values.

We also get information on k-anonymity of the combinations of these categorial variables: Currently, all observations breach 3- and 5-anonymity. There are only two persons in our dataset who share the same gender, age, education, plz, and years_in_job; therefore, our dataset also breaches 2-anonymity. 99% of participants are unique, meaning they have a unique combination on these demographic variables.

Upon closer inspection, these entries are IDs 94 and 122, who share many demographic details: both are 18 years old, live in Munich at postal code 80636, have been working their jobs for one year, and have a high school degree. If I knew my male, 18-year-old friend living at 80636 had participated in this study, I could not identify him based on a combination of thes data points alone. However, if I knew his job title, I could still identify him. Let's keep this in mind for further analyses.

```{r}
knitr::kable(data_withoutdirectidentifiers[data_withoutdirectidentifiers$id %in% c(94, 122), ])
```

This means that we currently do not reach any level of k-anonymity, since every other person has a unique combination.

For the **numerical key variables**, `sdcMicro` shows the disclosure risk. This parameter is a relative measure: it shows how much the parameters were changed after anonymization. Therefore, right now, `sdcMicro` does not know anything about the disclosure risk and makes the non-estimate that it is somewhere between 0 and 1.

**Other risk estimates**

`sdcMicro` offers even more risk estimates:

```{r}
# Risk summary
print(sdc_data, type = "risk")

# Individual risk
individualrisk <- get.sdcMicroObj(sdc_data, "risk")$individual

```

The individual risk shows a risk estimate for each individual in the dataset. `fk` is the frequency count in the sample and shows how many individuals share the same combination of key variables currently in our sdcObject. We can see that this is 1 for all participants except the IDs 94 and 122 I discussed above. `Fk` is the estimated frequency in the population; this only differs from `fk` in case of pre-defined sampling weights.
:::

At this point, you decide whether the level of privacy risks is acceptable. If so, you do not need anonymization but can skip to preparing your data for publication.

Save the sdcObject for further anonymization steps.

```{r}
saveRDS(sdc_data, "../sdc_micro.rds")
```

If the risk is not acceptable - as is the case here - we will start with the actual anonymization process. To that end, I will present you several techniques over the next few chapters.

#### Learning Objective

-   After completing this part of the tutorial, you will be able to calculate individual privacy risks for continuous variables by utilizing k-anonymity.

#### Exercises

-   calculate the privacy risk for certain variables

#### Resources, Links, Examples
