---
title: "Perturbative Techniques"
bibliography: ../references.bib
---

-   Not only masking of values, but distortion of values

-   goal: statistics run on the original dataset yield the same results as statistics run on the perturbed dataset

*Explain the idea more*

*Select most relevant techniques (swapping, noise?); potentially shorten or delete others*

*Add examples to each*

Techniques following @Carvalho2023_SurveyPrivacyPreservingTechniques

# Examples of Perturbative Techniques

## Swapping

-   exchanging values on certain variables between participants

-   types of swapping

    -   record swapping (also known as data swapping): for categorial variables; swapping the values on the variables (e.g., gender, country of residence); t-order equivalence means that the frequency tables for t variables are not changed (e.g., 1-order equivalence: same number of males and females as before, 2-order equivalence: same number of males and females from Switzerland and Germany respectively)

    -   rank swapping: for continuous variables; swapping values only within certain range of the rank to limit distortion of data

-   advantages

    -   removes relationship between record and individual

    -   can be used in one or more sensitive variables without disturbing the non-sensitive variables

    -   provides protection to rare and unique values

    -   not limited to the type of variable

-   disadvantages

    -   may produce number of cases with unusual combinations

    -   non-random swapping means work

    -   can severely distort statistics for subgroups

    -   not useful against attribute disclosures

## Re-sampling

-   idea: create averages of independent samples

-   bootstrap independent samples

-   use average of first sample for first row, then average of second sample for second row...

*check for correct understanding*

## Noise

-   also known as randomization

-   idea: add more or less random value (additive noise) or multiply by more or less random value (multiplicative noise)

-   noise can be correlated or uncorrelated with values

-   transformations after adding the noise are possible

-   differential privacy methods usually mean noise

::: {.callout-tip title="Differential Privacy"}
-   adds noise to data, leading to plausible deniability for any individual

-   results of analysis stay the same independent of noise

-   results of analysis stay the same, independent if one person is in there or not

-   diffpriv: An R Package for Easy Differential Privacy

-   "Even if the attacker already suspects X is the only possible HIV case in the dataset, the data release should not confirm or deny that suspicion.”
:::

## Microaggregation

-   idea: create groups of similar values and change these to an aggregate value (e.g., mean, median)

-   works better when groups are more homogeneous

## Rounding

-   round values to certain other values

## PRAM

-   Post RAndomisation Method
-   values on a categorical variable are recoded with a certain probability

## Shuffling

-   variation of swapping
-   generate new sensitive data based on similar distributional properties
-   change the order of sensitive values based on the rank of new sensitive data

# Keeping Utility

*Explain how to ensure that the statistics are the same (or reference utility section)*

# Pro and Contra of Using Perturbative Techniques

*Add pro and con list*

-   danger of reverse-engineering the perturbation technique applied

# Exercise

Perturbative techniques **modify** values rather than removing or generalising them. The data still looks complete and realistic, but individual values have been altered enough that re-identification becomes unreliable. Two common methods:

-   **Microaggregation** — records are grouped by similarity and values within each group are replaced by the group mean. Individual values are obscured but aggregate statistics are preserved.
-   **Adding noise** — small random amounts are added to numeric values. The distribution stays plausible but exact values are no longer trustworthy.

Both are available directly in `sdcMicro`.

## Exercise: Applying Perturbative Techniques

Continue working with `sdc_nonpert` from the previous exercise.

1.  **Apply microaggregation to `income`** using the default method (`"mdav"`). Use a group size of `k = 5`.

2.  **Add noise to `income`** as an alternative — apply additive noise with a noise level of `0.1` (10% of the standard deviation). Compare the result to microaggregation: which distorts the data more?

3.  **Compare the information loss** reported by `sdcMicro` after each step. Which method better balances risk and utility for this variable?

::: callout-tip
You cannot undo perturbation steps within the same `sdc` object. Create a fresh copy of `sdc_nonpert` before trying the second method so you can compare them side-by-side.
:::

::: callout-note
## Solution

### Step 1 – Microaggregation

```{r}
sdc_micro <- microaggregation(
  obj    = sdc_nonpert,
  variables = "income",
  aggr   = 5,          # group size k
  method = "mdav"      # Maximum Distance to Average Vector
)

print(sdc_micro, type = "numrisk")
```

`mdav` groups records by their distance to the group centroid, then replaces each value with the group mean. With `aggr = 5`, at least 5 records share the same income value, so singling out an individual is harder.

### Step 2 – Additive noise (on a fresh copy)

```{r}
sdc_noise <- addNoise(
  obj    = sdc_nonpert,
  variables = "income",
  noise  = 0.1         # noise level as fraction of SD
)

print(sdc_noise, type = "numrisk")
```

`addNoise()` draws from a normal distribution with mean 0 and standard deviation = `noise × sd(income)` and adds it to each value. Every record gets a unique (slightly wrong) income.

### Step 3 – Comparing information loss

```{r}
# Information loss after microaggregation
il_micro <- get.sdcMicroObj(sdc_micro, "utility")

# Information loss after noise addition
il_noise <- get.sdcMicroObj(sdc_noise, "utility")

il_micro
il_noise
```

**Interpretation:** Microaggregation typically produces lower IL1 (mean absolute deviation between original and perturbed values) because entire groups share one value — the average. Noise addition preserves individual-level variation better but introduces random error into every single record. For a dataset where income is used in regression analyses, noise addition is often preferable; for frequency tables or group comparisons, microaggregation is a safer choice.
:::

#### Learning Objective

-   After completing this part of the tutorial, you will be able to apply selected perturbative techniques in R.

#### Exercise

-   Apply one or two techniques for certain variables in R

#### Resources, Links, Examples

-   adding noise with the `sdcMicro` R package

# References
