---
title: "Perturbative Techniques"
bibliography: ../references.bib
---

```{r Setup}
#| include: false

library(sdcMicro)
library(tidyverse)

data_withoutidentifiers <- read.csv("../SimulatedData_noidentifiers.csv") # Change based on the location of your data file

sdc_nonpert <- readRDS("../sdc_nonpert.rds") # Change based on the location of your Sdc-object file
```

-   Not only masking of values, but distortion of values

-   goal: statistics run on the original dataset yield the same results as statistics run on the perturbed dataset

The core idea behind perturbative techniques is to introduce controlled changes to data values. Unlike non-perturbative techniques that hide or remove information, perturbative methods keep all records and variables in the dataset but alter the values themselves. The key requirement is that the changes should be large enough to prevent re-identification but small enough that the overall statistical properties of the data (means, variances, correlations) are preserved.

*Select most relevant techniques (swapping, noise?); potentially shorten or delete others*

*Add examples to each*

Techniques following @Carvalho2023_SurveyPrivacyPreservingTechniques

# Examples of Perturbative Techniques

## Swapping

-   exchanging values on certain variables between participants

-   types of swapping

    -   record swapping (also known as data swapping): for categorial variables; swapping the values on the variables (e.g., gender, country of residence); t-order equivalence means that the frequency tables for t variables are not changed (e.g., 1-order equivalence: same number of males and females as before, 2-order equivalence: same number of males and females from Switzerland and Germany respectively)

    -   rank swapping: for continuous variables; swapping values only within certain range of the rank to limit distortion of data

-   advantages

    -   removes relationship between record and individual

    -   can be used in one or more sensitive variables without disturbing the non-sensitive variables

    -   provides protection to rare and unique values

    -   not limited to the type of variable

-   disadvantages

    -   may produce number of cases with unusual combinations

    -   non-random swapping means work

    -   can severely distort statistics for subgroups

    -   not useful against attribute disclosures

## Re-sampling

-   idea: create averages of independent samples

-   bootstrap independent samples

-   use average of first sample for first row, then average of second sample for second row...

To be more precise: you draw multiple bootstrap samples from the original data, compute the statistic of interest (e.g., the mean) for each sample, and then replace original values with these bootstrapped estimates. The result preserves the overall distribution but individual values no longer correspond to real observations.

*check for correct understanding*

## Noise

-   also known as randomization

-   idea: add more or less random value (additive noise) or multiply by more or less random value (multiplicative noise)

-   noise can be correlated or uncorrelated with values

-   transformations after adding the noise are possible

-   differential privacy methods usually mean noise

::: {.callout-tip title="Differential Privacy"}
-   adds noise to data, leading to plausible deniability for any individual

-   results of analysis stay the same independent of noise

-   results of analysis stay the same, independent if one person is in there or not

-   diffpriv: An R Package for Easy Differential Privacy

-   "Even if the attacker already suspects X is the only possible HIV case in the dataset, the data release should not confirm or deny that suspicion.”
:::

## Microaggregation

-   idea: create groups of similar values and change these to an aggregate value (e.g., mean, median)

-   works better when groups are more homogeneous

## Rounding

-   round values to certain other values

## PRAM

-   Post RAndomisation Method
-   values on a categorical variable are recoded with a certain probability

## Shuffling

-   variation of swapping
-   generate new sensitive data based on similar distributional properties
-   change the order of sensitive values based on the rank of new sensitive data

# Keeping Utility

After applying any perturbative technique, you should compare key statistics (means, standard deviations, correlations, regression coefficients) between the original and perturbed datasets. If the differences are small enough for your purposes, the perturbation has preserved utility. We cover more formal utility measures in the chapter on [Balancing Utility and Privacy](../3_PRIVACY_OPENNESS/3_1_Balancing.qmd).

# Pro and Contra of Using Perturbative Techniques

Pro:

-   all records remain in the dataset (no deletion)

-   statistical properties can be preserved, flexible

-   ifferent noise levels or group sizes let you tune the privacy-utility trade-off

Con:

-   individual values are no longer accurate (not suitable if exact values matter)

-   risk of reverse-engineering the perturbation if the method is known

-   an distort subgroup statistics if applied too aggressively

# Exercise

Perturbative techniques **modify** values rather than removing or generalising them. The data still looks complete and realistic, but individual values have been altered enough that re-identification becomes unreliable. Two common methods:

-   **Microaggregation** — records are grouped by similarity and values within each group are replaced by the group mean. Individual values are obscured but aggregate statistics are preserved.
-   **Adding noise** — small random amounts are added to numeric values. The distribution stays plausible but exact values are no longer trustworthy.

Both are available directly in `sdcMicro`.

## Exercise: Applying Perturbative Techniques

Continue working with `sdc_nonpert` from the previous exercise.

1.  **Apply microaggregation to `income`** using the default method (`"mdav"`). Use a group size of `k = 5`.

2.  **Add noise to `income`** as an alternative — apply additive noise with a noise level of `0.1` (10% of the standard deviation). Compare the result to microaggregation: which distorts the data more?

3.  **Compare the information loss** reported by `sdcMicro` after each step. Which method better balances risk and utility for this variable?

::: callout-tip
You cannot undo perturbation steps within the same `sdc` object. Create a fresh copy of `sdc_nonpert` before trying the second method so you can compare them side-by-side.
:::

::: {.callout-note collapse="true"}
## Solution

### Step 1 – Microaggregation

```{r}
sdc_micro <- microaggregation(
  obj    = sdc_nonpert,
  variables = "income",
  aggr   = 5,          # group size k
  method = "mdav"      # Maximum Distance to Average Vector
)

print(sdc_micro, type = "numrisk")
```

`mdav` groups records by their distance to the group centroid, then replaces each value with the group mean. With `aggr = 5`, at least 5 records share the same income value, so singling out an individual is harder.

### Step 2 – Additive noise (on a fresh copy)

```{r}
sdc_noise <- addNoise(
  obj    = sdc_nonpert,
  variables = "income",
  noise  = 0.1         # noise level as fraction of SD
)

print(sdc_noise, type = "numrisk")
```

`addNoise()` draws from a normal distribution with mean 0 and standard deviation = `noise × sd(income)` and adds it to each value. Every record gets a unique (slightly wrong) income.

### Step 3 – Comparing information loss

```{r}
# Information loss after microaggregation
il_micro <- get.sdcMicroObj(sdc_micro, "utility")

# Information loss after noise addition
il_noise <- get.sdcMicroObj(sdc_noise, "utility")

il_micro
il_noise
```

**Interpretation:** Microaggregation typically produces lower IL1 (mean absolute deviation between original and perturbed values) because entire groups share one value — the average. Noise addition preserves individual-level variation better but introduces random error into every single record. For a dataset where income is used in regression analyses, noise addition is often preferable; for frequency tables or group comparisons, microaggregation is a safer choice.

Save objects

```{r}
saveRDS(sdc_micro, "../sdc_micro.rds")
saveRDS(sdc_noise, "../sdc_noise.rds")

```
:::

#### Learning Objective

-   After completing this part of the tutorial, you will be able to apply selected perturbative techniques in R.

#### Exercise

-   Apply one or two techniques for certain variables in R

#### Resources, Links, Examples

-   adding noise with the `sdcMicro` R package

# References
