---
title: "All Excercises (With Solutions)"
bibliography: ../references.bib
---

# Create Data For Exercises

Code for creating the dataset (not executed)

```{r}
#| label: data-generation
#| eval: false   # <-- This block is not executed
#| code-fold: true
#| code-summary: "Show code for data generation"

library(charlatan)
library(tidyverse)
library(writexl)

set.seed(111)


# Load vector with German postal codes
plz_list <- read_xlsx("../list_plz.xlsx") %>% 
  pull(PLZ) %>%
  as.numeric()

# Make munich plz occur more often 
munich_plz <- c("80331", "80333", "80539", "80636", "80799")
mp <- runif(5, 0.04, 0.07)

# German name generator
fake <- PersonProvider_de_DE$new()

# Some IPv4 ranges commonly allocated in Germany
ip_prefix <- c(
  "2.160",   # Deutsche Telekom
  "31.16",   # Vodafone
  "46.5",    # Hetzner / German hosting
  "78.48",   # Deutsche Telekom
  "79.192",  # Deutsche Telekom
  "84.128",  # Deutsche Telekom
  "87.128",  # Deutsche Telekom
  "91.0",    # various German ISPs
  "93.192",  # German ISP ranges
  "217.0"    # Deutsche Telekom legacy range
)

# Common email domains
domains <- c(
  "gmail.com","outlook.com","hotmail.com",
  "icloud.com","yahoo.com",
  "gmx.de","web.de",
  "bluewin.ch","t-online.de","vodafone.de"
)

people <- tibble(
  
  # ---- basic demographic variables ----
  id = 1:200,
  gender = factor(sample(
    c("male","female","non-binary"),
    size = 200,
    replace = TRUE,
    prob = c(0.48,0.48,0.04)
  )),
  age = sample(18:90, 200, replace = TRUE),
  income = round(rnorm(200, mean = 50000, sd = 20000), 2),
  plz = as.numeric(sample(c(munich_plz, plz_list), 200, replace = TRUE, prob = c(mp, rep((1 - sum(mp)) / length(plz_list), length(plz_list))))), 


  # ---- religion ----
  religion = factor(sample(
    c("Catholicism","Protestantism","Islam","Eastern Orthodoxy","Judaism", "Buddhism", "None"),
    200, replace = TRUE,
    prob = c(0.27,0.25,0.06,0.02,0.01,0.01,0.38)
  )),
  
  # ---- socio‑economic variables ----
  job_title = factor(ch_job(200)),
  education = factor(sample(
    c("no degree","trade school","high school",
      "university","doctoral title"),
    200, replace = TRUE, prob = c(0.05,0.20,0.50,0.20,0.05)
  )),
  
  # ---- political opinion items (Likert 1–5) ----
  pol_immigration = sample(1:5, 200, replace = TRUE),
  pol_environment = sample(1:5, 200, replace = TRUE),
  pol_redistribution = sample(1:5, 200, replace = TRUE),
  pol_eu_integration = sample(1:5, 200, replace = TRUE),
  
  # ---- work experience ----
  years_in_job = round(rexp(200, rate = 0.15))  # right-skewed: most people have few years, some have many
  
  ) %>%
  
  # ---- generate names depending on gender ----
  rowwise() %>%
  mutate(
    first_name = case_when(
      gender == "female" ~ fake$first_name_female(),
      gender == "male" ~ fake$first_name_male(),
      TRUE ~ fake$first_name()
    ),
    last_name = fake$last_name()
  ) %>%
  ungroup() %>%
  
  # ---- combine first and last name ----
  mutate(name = paste(first_name, last_name)) %>%
  
  # ---- generate realistic email addresses ----
  rowwise() %>%
  mutate(
    email = if (sample(c(TRUE,FALSE),1)) {
      # firstname.lastname@domain
      paste0(
        tolower(first_name), ".", tolower(last_name),
        "@", sample(domains,1)
      )
    } else {
      # firstname[randomnumber]@domain
      paste0(
        tolower(first_name), sample(10:99,1),
        "@", sample(domains,1)
      )
    }
  ) %>%
  ungroup() %>%
  
 # ---- generate plausible German IPv4 addresses ----
  mutate(
    ip_address = paste0(
      sample(ip_prefix, n(), replace = TRUE), ".",
      sample(0:255, n(), replace = TRUE), ".",
      sample(0:255, n(), replace = TRUE)
    )
  ) %>%
  
  # ---- remove helper columns ----
  select(id, name, email, plz, gender, age, income, years_in_job, religion, job_title, education, pol_immigration, pol_environment, pol_redistribution, pol_eu_integration, ip_address)

people

# Save as .xlsx
write_xlsx(people, "../SimulatedData.xlsx")
```

# 1_3 Personal Data

Here is a simulated dataset of 200 Germans. The data's purpose is to answer whether certain political opinions are linked to religion. This dataset will be used throughout the exercises.

Download the dataset here:

```{r}
#| label: download-link
#| echo: false

# Generates a clickable download link for the GitHub Pages website
xfun::embed_file(
  "../SimulatedData.xlsx",
  text = "⬇️ Download SimulatedData.xlsx"
)
```

Copy this to a new Markdown file in RStudio to import the data for further analysis:

```{r}
#| label: data-import
#| message: false

library(tidyverse)
library(readxl)

# Load data based on downloaded file
data <- read_xlsx("../SimulatedData.xlsx") # Change based on the location of your data file

data
```

::: callout-note
## Data Dictionary

| Variable Name | Description | Item | Values |
|------------------|------------------|------------------|------------------|
| id | Number assigned to each participant in order of participation | (assigned in background) | Integer; 1-200 |
| name | First and last name of participant | Please indicate your full name (first and last name) | String of characters |
| email | Email address of participant | What is your e-mail address? | String of characters |
| plz | German postal code | What is your postal code? | String of characters |
| gender | Gender of participant | What is your gender? | Factor; "male"/"female"/"non-binary" |
| age | Age of participant in years | What is your age in years? | Integer; 18-100 |
| income | Personal annual income in Euros | What was your income over the last twelve months | Integer |
| religion | Religion of participant | What is your religion? | Factor; "Catholicism","Protestantism","Islam","Eastern Orthodoxy","Judaism", "Buddhism", "Hinduism", "Other", "None" |
| job_title | Title of job of partcipant | What is your job title? | String of characters |
| education | Highest degree of education | What is your highest degree of education? | Factor; "no degree","trade school","high school", "university","doctoral title" |
| pol_immigration | Likert item measuring opinion on immigration | The government should limit immigration more strictly than it currently does. | Integer; 1-5 |
| pol_environment | Likert item measuring opinion on environment | Protecting the environment should be a top priority, even if it slows economic growth. | Integer; 1-5 |
| pol_redistribution | Likert item measuring opinion on redistribution of wealth | The government should reduce income differences between rich and poor. | Integer; 1-5 |
| pol_eu_integration | Likert item measuring opinion on membership in EU | Our country benefits from being a member of the European Union. | Integer; 1-5 |
| ip_address | IP address (version 4) of participant's device when answering survey | (collected in background) | String of characters |
| years_in_job | Number of years the participant has been in their current job | How many years have you been in your current job? | Integer; 0–n |
:::

## Exercise: Direct Identifiers, Indirect Identifiers, and Special Categories

Inspect the dataset `data`. Answer the following questions:

1\. Which columns contain direct identifiers?

2\. Which contains indirect identifiers?

3\. Which contain special categories of personal data?

4\. Is there any column that does not contain personal data?

::: callout-note
## Solution

### Direct identifiers: `name`, `email`, `ip_address`

Name is probably the most obvious identifier there is.

Email addresses are considered direct identifiers since they often contain names and therefore, clearly specify certain individuals. However, even when not containing names, they can act as indirect identifiers by linking to an account or revealing a person's organization.

IP addresses are often considered direct identifiers since they can be linked back to specific devices, and by using Internet service providers' data, to individuals. \[TODO\] Check

### Indirect identifiers: `gender`, `age`, `education`, `job_title`, `income`, `plz`, `years_in_job`

Gender, age, education, job title, postal code, income, and years in job are indirect identifiers, since they cannot lead to idenitification of an individual on their own but may be linked with each other or external knowledge. When an attacker knows of a person who participated in the study and knows their job title, they could identify the person in the data.

### Special category data: all `pol\_` variables (political opinion), `religion`

Explanation

### Personal data: All columns contain personal data.

Any data that can be linked to an individual is considered personal data. As long as identification (e.g., via direct identifiers such as `name`) is possible, it is considered personal data and GDPR applies.
:::

Delete all direct identifiers from the dataset.

::: callout-note
## Solution

Any solution that deletes the name, email address, and IP address is correct. I use tidyverse syntax:

```{r}

data_withoutdirectidentifiers <- data %>% 
  select(-name, -email, -ip_address)
  
```
:::

# 2_1 Planning Privacy

## Using `sdcMicro` To Assess Privacy Risks

Let's start by creating a sdcMicro object using the class `sdcMicroObj`. The object becomes the working environment for all subsequent anonymization and risk assessment steps. It stores

-   the original dataset,
-   the roles of variables within the dataset,
-   risk measures,
-   anonymized versions of the variables,
-   utility measures, and
-   the history of modifications.

`sdcMicro` updates risk and utilty estimates after each anonymization step.

For this object, we start with defining the following:

-   `dat`: the dataset we are working with

-   `keyVars`: indirect identifiers that are categorial (i.e., have categories) we want to assess the risk of (e.g., age, gender)

-   `numVars`: continuous (i.e., numeric) indirect identifiers (e.g., income)

`sdcMicro` needs, at the least, the data and categorial indirect identifiers to assess risks. We can later add more information to the object.

## Exercise: Assessing Privacy Risks With `sdcMicro`

1.  Create an object within sdcMicro. Inspect the object.
2.  What k-anonymity level is currently reached?

Solution:

```{r}
library(sdcMicro)

sdc_data <- createSdcObj(
  dat = data_withoutdirectidentifiers, 
  keyVars = c("gender", "age", "education", "plz"),
  numVars = "income"
  )  

# Overall summary
sdc_data

# Risks summary
print(sdc_data, type = "risk")
# Individual risk
individualrisk <- get.sdcMicroObj(sdc_data, "risk")$individual # What is fk
# Continuous risk
print(sdc_data, type = "numrisk") # Same as overall summary
```

# 2_3 Applying Non-Perturbative Techniques With R

-   but: when publishing anonymization script: needs to be anonymous as well (e.g., not naming specific countries; see [Chapter DOCUMENTATION](../3_PRIVACY_OPENNESS/3_2_Documentation))

## Exercise: Applying Non-Perturbative Techniques

*TODO Decide on variables*

Apply non-perturbative techniques to the data to reduce re-identification risk.

The following `sdcMicro` functions are available to you:

-   `globalRecode()` — recodes a key variable into broader groups (e.g., exact age → age band)
-   `topBotCoding()` — caps extreme values at a given threshold
-   `localSuppression()` — suppresses individual cells in records that fall below a chosen k-anonymity threshold. I would advise against using this function since it does not report changes transparently. Instead, you can recode data manually, e.g., by using `mutate`, or by using a function such as `fct_lump_min()` from `forcats`. This function lumps all categories below a certain group size together.

Look at the key variables in `sdc_data` (`gender`, `age`, `education`, `plz`) and decide which technique is appropriate for each. There is no single correct solution — choose thresholds and groupings that make sense given the data, and be prepared to justify your choices. After applying your techniques, compare the new risk summary to the one from the previous exercise.

::: callout-tip
The `sdcMicro` functions operate directly on the `sdc` object and update risk estimates automatically. Check `?globalRecode`, `?topBotCoding`, and `?localSuppression` for argument details.

The `forcat` function `fct_lump_min()` is applied to the dataset itself. The `sdc` object needs to be updated afterwards.
:::

::: callout-tip
## Solution

One possible approach:

```{r}
# Recode age into bands
sdc_nonpert <- globalRecode(
  obj    = sdc_data,
  column = "age",
  breaks = c(17, 29, 44, 59, 74, Inf),
  labels = c("18-29", "30-44", "45-59", "60-74", "75+")
)  

print(sdc_nonpert, type = "risk") # check risk after making the change

# Generalize PLZ
sdc_nonpert <- globalRecode(
  obj    = sdc_data,
  column = "plz",
  breaks = c(0, 19999, 39999, 59999, 79999, 99999),
  labels = c("00–19", "20–39", "40–59", "60–79", "80–99")
)

# `globalRecode()` updates the key variable inside the `sdc` object directly and recalculates risk. The five regions map loosely to north, north-central, central, south-central, and south Germany.

# Top-code income
sdc_nonpert <- topBotCoding(
  obj         = sdc_nonpert,
  column      = "income",
  value       = quantile(data_withoutdirectidentifiers$income, 0.95),
  replacement = quantile(data_withoutdirectidentifiers$income, 0.95),
  kind        = "top"
)

print(sdc_nonpert, type = "risk") # check risk after making the change

# Top code years in job
# Inspect the distribution first to pick a sensible threshold
summary(data_withoutdirectidentifiers$years_in_job)

sdc_nonpert <- topBotCoding(
  obj         = sdc_nonpert,
  column      = "years_in_job",
  value       = 20,        # values above 20 years get capped
  replacement = 20,
  kind        = "top"
)

print(sdc_nonpert, type = "risk")
# `years_in_job` follows an exponential distribution, so a small number of people will have very high values. Capping at 20 years is a reasonable choice — adjust based on what the actual distribution looks like in your dataset.

# Apply local suppression to reach k = 3
sdc_nonpert <- localSuppression(sdc_nonpert, k = 3)  # Compare risk before and after

print(sdc_nonpert, type = "risk")

# Collapse rare religions
library(forcats)

# Inspect counts first
table(data_withoutdirectidentifiers$religion)

data_religion_recoded <- data_withoutdirectidentifiers %>%
  mutate(religion = fct_lump_min(religion, min = 10, other_level = "Other"))

table(data_religion_recoded$religion)

# Rebuild the sdc object with the recoded religion column
sdc_nonpert <- createSdcObj(
  dat     = data_religion_recoded,
  keyVars = c("gender", "age", "education", "plz"),
  numVars = c("income", "years_in_job")
)

# Now apply the previous steps again on the fresh object, then check
print(sdc_nonpert, type = "risk")
```

`fct_lump_min(religion, min = 10)` merges any category with fewer than 10 observations into `"Other"`. This is transparent — you can see exactly which categories were affected by comparing the `table()` outputs before and after. Groups like "Judaism" or "Buddhism" which have very few members in this dataset are natural candidates for collapsing, since a person could potentially be identified through their combination of rare religion and other attributes. *Compare to local suppression using function*
:::

# 2_4 Applying Perturbative Techniques With R

Perturbative techniques **modify** values rather than removing or generalising them. The data still looks complete and realistic, but individual values have been altered enough that re-identification becomes unreliable. Two common methods:

-   **Microaggregation** — records are grouped by similarity and values within each group are replaced by the group mean. Individual values are obscured but aggregate statistics are preserved.
-   **Adding noise** — small random amounts are added to numeric values. The distribution stays plausible but exact values are no longer trustworthy.

Both are available directly in `sdcMicro`.

## Exercise: Applying Perturbative Techniques

Continue working with `sdc_nonpert` from the previous exercise.

1.  **Apply microaggregation to `income`** using the default method (`"mdav"`). Use a group size of `k = 5`.

2.  **Add noise to `income`** as an alternative — apply additive noise with a noise level of `0.1` (10% of the standard deviation). Compare the result to microaggregation: which distorts the data more?

3.  **Compare the information loss** reported by `sdcMicro` after each step. Which method better balances risk and utility for this variable?

::: callout-tip
You cannot undo perturbation steps within the same `sdc` object. Create a fresh copy of `sdc_nonpert` before trying the second method so you can compare them side-by-side.
:::

::: callout-note
## Solution

### Step 1 – Microaggregation

```{r}
sdc_micro <- microaggregation(
  obj    = sdc_nonpert,
  variables = "income",
  aggr   = 5,          # group size k
  method = "mdav"      # Maximum Distance to Average Vector
)

print(sdc_micro, type = "numrisk")
```

`mdav` groups records by their distance to the group centroid, then replaces each value with the group mean. With `aggr = 5`, at least 5 records share the same income value, so singling out an individual is harder.

### Step 2 – Additive noise (on a fresh copy)

```{r}
sdc_noise <- addNoise(
  obj    = sdc_nonpert,
  variables = "income",
  noise  = 0.1         # noise level as fraction of SD
)

print(sdc_noise, type = "numrisk")
```

`addNoise()` draws from a normal distribution with mean 0 and standard deviation = `noise × sd(income)` and adds it to each value. Every record gets a unique (slightly wrong) income.

### Step 3 – Comparing information loss

```{r}
# Information loss after microaggregation
il_micro <- get.sdcMicroObj(sdc_micro, "utility")

# Information loss after noise addition
il_noise <- get.sdcMicroObj(sdc_noise, "utility")

il_micro
il_noise
```

**Interpretation:** Microaggregation typically produces lower IL1 (mean absolute deviation between original and perturbed values) because entire groups share one value — the average. Noise addition preserves individual-level variation better but introduces random error into every single record. For a dataset where income is used in regression analyses, noise addition is often preferable; for frequency tables or group comparisons, microaggregation is a safer choice.
:::

# 3_1 Balancing

Evaluate the utility of the dataset using sdcMicro functions

After anonymisation, you need to check that the data is still **useful** — that the patterns you care about (e.g., the relationship between religion and political opinion) have not been destroyed. `sdcMicro` provides utility measures, but it is also worth comparing distributions and running the actual analysis on both the original and anonymised data.

Key utility concepts:

-   **IL1** — average absolute difference between original and anonymised numeric values (lower = better)
-   **Eigen** — compares covariance structure; values close to 1 mean the relationships between variables are preserved
-   **`utility()` / `print(..., type = "utility")`** — overall summary
-   **Visual comparison** — plotting distributions side-by-side is often the most intuitive check

## Exercise: Evaluating Data Utility

Use `sdc_micro` (the microaggregation result from the previous exercise) and compare it to the original data.

1.  **Print the utility summary** of `sdc_micro`. What do IL1 and the eigen measure tell you?

2.  **Compare the income distribution** before and after anonymisation. Plot histograms of the original and anonymised income side-by-side.

3.  **Re-run the analysis that motivated data collection:** compute the mean political opinion scores (`pol_immigration`, `pol_redistribution`) by religion group on both the original and anonymised data. Are the group differences preserved?

4.  **Reflect:** based on your findings, is the anonymised dataset fit for purpose? Would you adjust any of the anonymisation steps?

::: callout-note
## Solution

### Step 1 – Utility summary

```{r}
print(sdc_micro, type = "utility")
```

IL1 close to 0 indicates little numeric distortion. Eigen values close to 1 indicate that the covariance structure between numeric variables is preserved. If you applied strong microaggregation (small `k` → large groups), you will see higher IL1 and eigen values further from 1.

### Step 2 – Compare income distributions

```{r}
library(ggplot2)

# Extract anonymised income from the sdc object
income_original   <- data_nonpert$income
income_anonymised <- get.sdcMicroObj(sdc_micro, "manipNumVars")$income

income_compare <- tibble(
  income = c(income_original, income_anonymised),
  version = rep(c("Original", "Anonymised"), each = length(income_original))
)

ggplot(income_compare, aes(x = income, fill = version)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  scale_fill_manual(values = c("Original" = "#2E4057", "Anonymised" = "#E07A5F")) +
  labs(
    title = "Income distribution: original vs. anonymised",
    x = "Annual income (€)", y = "Count", fill = NULL
  ) +
  theme_minimal()
```

The shapes should look similar. A heavily chunked distribution (tall bars at regular intervals) would indicate that microaggregation groups were too large.

### Step 3 – Check whether the key analysis still works

```{r}

# Analysis on original data
original_summary <- data_withoutdirectidentifiers %>%
  group_by(religion) %>%
  summarise(
    mean_immigration    = mean(pol_immigration,    na.rm = TRUE),
    mean_redistribution = mean(pol_redistribution, na.rm = TRUE),
    n = n()
  )

# Rebuild anonymised dataset with the manipulated variables
anon_data <- get.sdcMicroObj(sdc_micro, "manipKeyVars") %>%
  bind_cols(get.sdcMicroObj(sdc_micro, "manipNumVars")) %>%
  bind_cols(
    data_nonpert %>% select(religion, starts_with("pol_"))
  )

anonymised_summary <- anon_data %>%
  group_by(religion) %>%
  summarise(
    mean_immigration    = mean(pol_immigration,    na.rm = TRUE),
    mean_redistribution = mean(pol_redistribution, na.rm = TRUE),
    n = n()
  )

original_summary
anonymised_summary
```

If the group means are close between both summaries, the anonymisation has preserved the analytical signal. Political opinion variables were not altered (they are neither key variables nor numeric variables in the sdcMicro object), so differences here would only arise if the grouping structure changed dramatically.

### Step 4 – Reflection

There is no single correct answer. Consider:

-   If income distributions look very similar and group means are preserved → the anonymisation is fit for purpose for this analysis
-   If you used `k = 5` and results still look good, increasing to `k = 3` would reduce distortion while still providing some protection — but weigh this against the risk reduction you found in the earlier exercises
-   Variables like `plz_region` and `age_group` were generalised; if a future analysis requires finer geographic or age breakdowns, you would need to negotiate a different anonymisation strategy with the data controller
:::

# Documentation

Make changes to the data dictionary in accordance with the anonymization steps you made in this tutorial.

-   `extractManipData(sdc)`\
    Extracts the anonymized dataset.

-   `undolast(sdc)`\
    Reverts the last anonymization step.

-   `get.sdcMicroObj()`\
    Accesses internal components of the object.
