---
title: "Privacy Concepts"
bibliography: ../references.bib
---

# Privacy Risks

Disclosure risks according to @Carvalho2023_SurveyPrivacyPreservingTechniques

*Explain each risk more*

*Add research related examples to each risk*

-   **identity disclosure**: when an intruder can recognize that a record in the released dataset concerns an individual by matching QI values

-   **attribute disclosure**: when an intruder is able to determine new characteristics of a data subject based on the information available in the released data

-   **inferential disclosure**: when an intruder can infer the data subject's private information with high confidence from statistical properties of the released data

    -    ChatGPT example: 90% of females aged 40-50 have burnout --\> very likely, that person I know is in that group in the study has burnout; also might relate to people who are not part of that dataset

-   **membership disclosure**: when an intruder is able to conclude whether the private information regarding a certain individual is present or not in the dataset

    -   ChatGPT example: a dataset on unregistered immigrants was released. I can infer that one row belongs to my colleague --\> I know that my colleague is not registered

# Differential Privacy

Let's return to one of the " Not Clear" Cases from the exercises before, for example this one:

Customer reviews are published without names or emails, but the full free-text is kept. Some reviews contain details like *“As a doctor at Klinikum Grosshadern…”* or *“My husband, John, surprised me with this gift.”*

You might have realized that the amount of anonymisation or pseudonymisation here is not binary, but more a gradient dependent on the situation, the process and the algorithm used to

The headers below cause the table of contents/dropdown on the right to be visible.

References to external resources on topics like Data Security Fundamentals, Risk Management, and Individual Rights

# K-anonymity

Definition and Initial Purpose K-anonymity was developed by computer scientist Latanya Sweeney following her demonstration of how she could re-identify supposedly anonymized medical records released by the State of Massachusetts. By linking public medical records with voter registration records using quasi-identifiers like age, zip code, and gender, she was able to single out individuals, including the governor's medical records (p 54, @jarmul2023practical). The principle of k-anonymity attempts to mitigate such linkage attacks by ensuring that in any released dataset, each record is indistinguishable from at least k-1 other records concerning a set of "quasi-identifiers" (attributes that, when combined, can uniquely identify an individual). This means grouping people with similar sensitive attributes, and not releasing groups that have fewer than k people.

An example of K-anonnymity is given below.

*Move package info to next part*

# The sdcMicro R package

sdcMicro is a R package specialized on the anonymisation of micro (small) data sets. sdcMicro offers a command line and a Graphical User interface (GUI).

Let's generate some fake data:

```{r}
data <- data.frame(
  age = c(34, 35, 36, 36, 36, 40, 40, 41, 42, 42),
  gender = c("M", "M", "M", "F", "F", "M", "F", "F", "M", "M"),
  zipcode = c("12345", "12345", "12345", "12345", "12345", "67890", "67890", "67890", "67890", "67890"),
  diagnosis = c("Flu", "Cold", "Flu", "Cold", "Allergy", "HIV", "HIV", "Flu", "Cold", "Allergy")
)

```

```{r}
#| echo: false
knitr::kable(
  head(data),
  caption = "Raw data example"
)
```

# References
